---
title: "Choosing an allocation procedure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an allocation procedure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzalloc)
set.seed(20260908)
```

Every procedure in this package answers the same question, one subject
at a time: which arm does the next person go to? They differ in what
they are willing to give up to get it. This vignette is about that
trade.

## The trade

Two properties are in tension.

**Balance.** Unequal arm sizes cost power, and an imbalance that falls
along a prognostic covariate costs more than that, because it
confounds the treatment effect with the covariate.

**Unpredictability.** If an investigator can guess the next assignment,
they can choose whom to enrol next, and selection bias enters a trial
that was supposed to be randomized. This is not hypothetical: it is why
allocation concealment is a standard item in risk-of-bias assessment.

Nothing achieves both perfectly. Simple randomization is maximally
unpredictable and does not bound imbalance at all; forcing exact
balance makes the tail of every block deterministic.

```{r}
n <- 100
imbalance_path <- function(x) cumsum(2 * x - 1)

simple <- alloc_simple(n)
blocked <- alloc_permuted_block(n, block_size = 4)

c(simple = max(abs(imbalance_path(simple))),
  blocked = max(abs(imbalance_path(blocked))))
```

Simple randomization wanders; the block design cannot exceed half a
block. The cost is that in a block of four, once three assignments are
known the fourth is determined.

## Bounding imbalance without tilting every coin

The biased-coin family tilts the coin at every step. Efron's design
sends the next subject to the smaller arm with probability `p`:

```{r}
efron <- alloc_efron(n, p = 2 / 3)
max(abs(imbalance_path(efron)))
```

The big stick does something different: it leaves the coin fair until
the imbalance actually reaches a boundary, and only then forces the
deficient arm. It pays for the bound only when the bound binds.

```{r}
stick <- alloc_big_stick(n, boundary = 3)
max(abs(imbalance_path(stick)))
```

The maximal procedure goes further. Among all sequences that respect a
boundary *and* finish in balance, it draws uniformly, which makes it
the most random procedure subject to those constraints.

```{r}
mx <- alloc_maximal(20, boundary = 2)
c(final_imbalance = sum(2 * mx - 1),
  max_excursion = max(abs(imbalance_path(mx))))
```

Terminal balance is part of its definition, not an accident, so the
sample size must be even.

## Balancing covariates

Bounding the total arm sizes says nothing about whether the arms are
comparable on prognostic factors. Two approaches address that.

**Stratification** runs a separate block sequence within each cell of
the covariate cross-classification. It gives the strongest balance when
the cells are well populated, and degrades when they are not: the
number of cells is the product of the numbers of levels, so with a few
factors most cells hold a handful of subjects and each contributes its
own partial block.

**Minimization** balances the margins instead, which is what lets it
handle many factors at once.

```{r}
cv <- data.frame(
  sex = sample(c("F", "M"), 200, TRUE),
  stage = sample(c("I", "II", "III"), 200, TRUE),
  site = sample(c("a", "b", "c", "d"), 200, TRUE)
)

worst_marginal <- function(a) allocation_imbalance(a, cv)$max_marginal

c(simple = worst_marginal(alloc_simple(nrow(cv))),
  stratified = worst_marginal(alloc_stratified_block(cv)),
  minimization = worst_marginal(alloc_pocock_simon(cv)))
```

With three factors and 24 cells, minimization holds the margins tighter
than stratification does.

### Weighting factors that matter more

Covariates are not equally prognostic, and minimization is the only
member of this family that can say so. A stratified design can include
a factor or exclude it; it cannot weight it.

```{r}
even <- alloc_pocock_simon(cv, weights = c(sex = 1, stage = 1, site = 1))
tilted <- alloc_pocock_simon(cv, weights = c(sex = 1, stage = 8, site = 1))

sapply(list(even = even, tilted = tilted), function(a) {
  m <- allocation_imbalance(a, cv)$marginal
  c(stage = max(abs(m$stage)), sex = max(abs(m$sex)))
})
```

Up-weighting `stage` improves stage balance and costs sex balance. The
weighting shifts effort; it does not add any.

## Judging the result on the scale that matters

Counting covariate imbalance treats every factor as equally important.
It is not. `weighted_imbalance()` weights each covariate's imbalance by
its association with the outcome, which is the scale on which an
imbalance actually costs you something.

```{r}
cvn <- data.frame(
  strong = sample(0:1, 200, TRUE),
  weak = sample(0:1, 200, TRUE)
)
a <- alloc_simple(200)

# 'strong' predicts the outcome five times as hard as 'weak'
weighted_imbalance(a, cvn, beta = c(strong = 1.0, weak = 0.2))
```

A design can look well balanced on counts while being materially
unbalanced on this measure, if the imbalance happens to land in the
covariate that predicts the outcome. The converse also holds, which is
the more comforting case.

## One interface

Every procedure is reachable through `allocate()`, so a simulation can
iterate over procedures without a `switch` at each call site.

```{r}
allocation_schemes()
```

```{r}
sapply(c("simple", "efron", "big_stick", "maximal"),
       function(s) max(abs(imbalance_path(allocate(s, n = 40)))))
```

## Where to look next

`vignette("covariate-adaptive")` covers the minimization family in
more detail, including the joint-balance and minimal-sufficient-balance
variants and the response-adaptive design.
