---
title: "Covariate-adaptive randomization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate-adaptive randomization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zzalloc)
set.seed(20260908)
cv <- data.frame(
  sex = sample(c("F", "M"), 240, TRUE),
  stage = sample(c("I", "II", "III"), 240, TRUE)
)
```

The procedures here all look at the covariates of the arriving subject
before assigning them. They differ in what they consider imbalanced and
in how often they are willing to intervene.

## Pocock-Simon minimization

For each arriving subject the procedure computes what the imbalance
would be under each assignment, summed over the levels that subject
actually has, and assigns to the arm minimizing that total with
probability `p`.

```{r}
a <- alloc_pocock_simon(cv, p = 0.8)
allocation_imbalance(a, cv)$marginal
```

Only the subject's own levels enter the score. Levels they do not
belong to are unaffected by where this subject goes, and would
contribute the same constant to both hypothetical totals.

`p` controls how hard the procedure pushes. At `p = 0.5` the score
never influences the draw and the procedure is simple randomization; at
`p = 1` it is deterministic minimization, which is Taves's original
proposal and the most predictable member of the family.

```{r}
mean_imbalance <- function(p, reps = 40) {
  mean(replicate(reps,
    allocation_imbalance(alloc_pocock_simon(cv, p = p), cv)$max_marginal))
}
sapply(c(0.5, 0.7, 0.9, 1.0), mean_imbalance)
```

### The measure argument

With two arms, `"sd"` and `"diff"` give identical allocations: the
standard deviation of two counts is their absolute difference divided
by the square root of two, and the constant cancels in the comparison.
Only `"squared"` can order candidate assignments differently, because
a tie in the sum of absolute differences is not a tie in the sum of
squares: `1 + 3` equals `2 + 2`, but `1 + 9` does not equal `4 + 4`.

```{r}
set.seed(1); m_sd <- alloc_pocock_simon(cv, measure = "sd")
set.seed(1); m_df <- alloc_pocock_simon(cv, measure = "diff")
identical(m_sd, m_df)
```

## Hu-Hu joint balance

Marginal balance can hide imbalance inside cells, which matters when
the outcome depends on a covariate interaction. `alloc_hu_hu()` scores
three things at once: overall imbalance, marginal imbalance, and
imbalance within the subject's own joint stratum.

```{r}
worst_stratum <- function(a) max(abs(allocation_imbalance(a, cv)$stratum))
c(pocock_simon = worst_stratum(alloc_pocock_simon(cv)),
  hu_hu = worst_stratum(alloc_hu_hu(cv, stratum_weight = 2)))
```

Setting both extra weights to zero recovers Pocock-Simon exactly:

```{r}
set.seed(2); ps <- alloc_pocock_simon(cv, p = 0.8)
set.seed(2); hh <- alloc_hu_hu(cv, p = 0.8, overall_weight = 0,
                               stratum_weight = 0)
identical(ps, hh)
```

## Minimal sufficient balance

The objection to minimization is that it tilts the coin at every
allocation, to prevent imbalances that mostly would not have occurred.
Minimal sufficient balance intervenes only when an imbalance has
actually appeared: each covariate is tested with a chi-square test on
its arm-by-level table, and only a covariate whose p-value falls below
`threshold` casts a vote.

```{r}
msb <- alloc_msb(cv, threshold = 0.3)
mean(msb)  # overall allocation is left alone
```

Because it targets the covariate *distribution* across arms rather than
the count difference within each level, it improves on simple
randomization by a smaller margin on per-level counts than a per-level
rule would, and by a similar margin on the distributional scale it
actually targets.

```{r}
prop_gap <- function(a) {
  tb <- table(factor(a, levels = c(0, 1)), cv$sex)
  abs(tb[2, 1] / sum(tb[2, ]) - tb[1, 1] / sum(tb[1, ]))
}
c(msb = mean(replicate(40, prop_gap(alloc_msb(cv)))),
  simple = mean(replicate(40, prop_gap(alloc_simple(nrow(cv))))))
```

## Response-adaptive allocation

`alloc_cara()` is different in kind from everything above. Those
procedures balance covariates and are agnostic to outcomes; this one
deliberately induces imbalance *in response* to outcomes, sending
subjects preferentially to the arm that is working better for people
like them. The objective is ethical rather than statistical.

```{r}
set.seed(3)
cvc <- data.frame(x = rnorm(150))
rf <- function(trt, row) rbinom(1, 1, plogis(-0.2 + 1.5 * trt + 0.5 * row$x))
out <- alloc_cara(cvc, rf, burn_in = 40)
c(burn_in = mean(out$trt[1:40]), adapted = mean(out$trt[41:150]))
```

Two cautions. The assignment sequence depends on the responses, so
standard tests require more care than under covariate-adaptive
randomization alone. And the implementation assumes responses arrive
before the next subject is randomized, which is frequently false:
with delayed responses the procedure behaves closer to its burn-in.

By default it draws fresh parameters per imputation, propagating
estimation uncertainty. Setting `param_uncertainty = FALSE` conditions
on the fitted model instead, which understates that uncertainty.
